A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-15 paper.tex
Fixup intro

\section*{Introduction}
% Inferring the genetic relationships between sampled genomes in the form of a
% phylogenetic tree is a necessary prerequisite for many analyses of
% species~\citep{rannala2003genetics}, non-recombining sections of
% DNA~\citep{cann1987mitochondrial,underhill2001annalsofhumangenetics}
% and viruses~\citep{grenfell2004science}.
% A phylogenetic tree captures information about the genealogy
% of the sample in an elegant and concise
% way, by postulating a set of ancestors common to the samples
% together with paths of inheritance between them. A rich literature exists
% on analysing the mathematical properties of these
% trees~\citep{steel2016phylogeny}, and numerous
% methods exist to infer them~\citep{felsenstein2004inferring}.
% However, the genetic genealogy of a recombining organism
% cannot be captured by a single tree: instead,
% the inheritance of each section of the genome may be described
% by a (slightly) different local tree~\citep{hudson1983properties}.
% Inferring complex
% genealogies of this nature
% presents profound technical difficulties, meaning genealogy-based approaches have
% not been widely used in population and statistical genetics.
% Until recently, analysis of
% population genetic data has therefore primarily focussed on summaries derived
% from the local trees~\citep{tajima1983evolutionary,tavare1984line}
% rather than using the full underlying genealogy itself. Such
% derived measures include allele frequencies and site frequency
% spectra~\citep{achaz2009frequency,ralph2020efficiently},
% patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and
% principal components~\citep{mcvean2009genealogical}.
Estimating the detailed genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using these ancestries as inputs have begun to
appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
nowbandegani2023extremely}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

% TODO need to mention ARG-as-a-data-structure to clarify.
% What is the problem?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}.
Originally, the ARG was based on a stochastic
process~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}
which modelled the coalescent with recombination \citep{hudson1983properties}.
Subsequently, it has come instead to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ a means of
describing the sequence of correlated local trees that
are the \emph{outcome} of some such ancestral process.
However, the distinction between process and structure is not clear cut
and subfields use the term
differently (see Appendix XXX).
%% Sentence deleted: point made already
% ARG inference was for many years synonymous [FIXME: this isn't really true,
% what about parsimony ARGs?]
% with estimating realisations
% of the coalescent with recombination (with, it must be said, limited success),
% and the distinction between process and realisation was not especially important.

@@

The word ``ARG'' is now widely used to refer to genetic genealogies in
general~\citep[MORE][]{mathieson2020ancestry}, including
varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
However, this broad usage means there is disagreement on the
exact object being defined. This causes problems, for example,
not just when evaluating inferred ARGs, but even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer}.
Perhaps most importantly, some definitions of ARGs
(especially those linked to a specific stochastic process) are overly narrow, which
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we
assess success, and the expected limitations on the precision
and accuracy of our inferences.

% What do we do?
In this paper we attempt to address these problems by
providing a precise definition of an ARG that is general enough to encompass
the output of all popular simulation and inference methods,
yet simple enough to be encoded via a conventional mathematical graph.
%% Yan: I added the above as both the ARGON and the Shipilova definitions
%% use non-conventional graph structures for their encoding (e.g. hypergraphs, weird edges, etc)
We follow the well established principle that it is advantageous
to decouple the presence of mutations from the genealogical structure
itself \citep[contra][chap.~3]{gusfield2014recombinatorics}.
We propose a definition of genetic genealogy purely in terms of specific genomes
and their intervals of genetic inheritance, and refer to this as
a ``genome ARG'', or ``gARG''.
This contrasts with the classical Griffiths approach,
which is based on recording evolutionary events, and which we call
an ``event ARG'', or ``eARG''.
We show that the gARG encoding has a range of advantages over the
eARG approach, being both flexible and efficient.
In particular, we argue that the goal of inferring fully precise
ARGs---a necessity under the eARG encoding---should be
% weird phrasing for a criticism?
critically examined
%
as part of the process of fully decoupling our ARG data
structure from any specific generative model.

@@

\section*{Event ARGs}\label{eARG}
In keeping with its original derivation in terms of a backward-time stochastic
process (see Appendix~\ref{sec-arg-history}), the classical definition of an ARG data structure is in terms of events, and we therefore refer to it as an ``event ARG'' (eARG).
In an eARG, nodes represent
common ancestor or recombination events that occurred in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage with one parent, whereas in a recombination
event a lineage is split into two independent
ancestral lineages, and a list of two parents is required.
As illustrated in Figure~\ref{fig-arg-data-structure},
the regions of the genome which are transmitted are usually
specified by associating recombination nodes with a breakpoint.
In other words, eARGs traditionally
carry positional information on their nodes, rather than
storing genome positions as edge annotations.

@@

% \section*{Inheritance and ancestral material}\label{Ancestry_resolution}
\section*{Prospective vs retrospective ARGs}
ARGs are usually thought of as a retrospective structure, in which we
reason about the genomes that are genetic ancestors of some set of
sample genomes. Figure~\ref{fig-arg-in-pedigree}
is retrospective because it only includes information
about genetic inheritance relative to the sample individuals
$I_1$ and $I_2$ (genomes \textsf{a}--\textsf{d}).
Genetic inheritance must have occurred between (e.g.)
$I_8$ and her daughter $I_6$, but this is not recorded because
% this is probably not true, given we know the complete pedigree?
we have no information about it, given these samples.
Event ARGs, given their roots in the coalescent with recombination,
are inherently retrospective. They define the evolutionary
events that have occurred in the genetic ancestry of a sample.
% This is weak, rephrase.
An event ARG must have some set of samples, or there
will be nothing to define events in terms of.

>>>>

\section*{Introduction}
% Inferring the genetic relationships between sampled genomes in the form of a
% phylogenetic tree is a necessary prerequisite for many analyses of
% species~\citep{rannala2003genetics}, non-recombining sections of
% DNA~\citep{cann1987mitochondrial,underhill2001annalsofhumangenetics}
% and viruses~\citep{grenfell2004science}.
% A phylogenetic tree captures information about the genealogy
% of the sample in an elegant and concise
% way, by postulating a set of ancestors common to the samples
% together with paths of inheritance between them. A rich literature exists
% on analysing the mathematical properties of these
% trees~\citep{steel2016phylogeny}, and numerous
% methods exist to infer them~\citep{felsenstein2004inferring}.
% However, the genetic genealogy of a recombining organism
% cannot be captured by a single tree: instead,
% the inheritance of each section of the genome may be described
% by a (slightly) different local tree~\citep{hudson1983properties}.
% Inferring complex
% genealogies of this nature
% presents profound technical difficulties, meaning genealogy-based approaches have
% not been widely used in population and statistical genetics.
% Until recently, analysis of
% population genetic data has therefore primarily focussed on summaries derived
% from the local trees~\citep{tajima1983evolutionary,tavare1984line}
% rather than using the full underlying genealogy itself. Such
% derived measures include allele frequencies and site frequency
% spectra~\citep{achaz2009frequency,ralph2020efficiently},
% patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and
% principal components~\citep{mcvean2009genealogical}.
Estimating the genetic genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
nowbandegani2023extremely}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

% TODO need to mention ARG-as-a-data-structure to clarify.
% What is the problem?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}.
Originally, ARGs were defined by Griffiths and colleagues as an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised as a
graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Subsequently, ARGs have come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
the \emph{outcome} of such a random process or an inference
of the genetic genealogy of a sample of genome sequences.
However, the distinction between process and structure is not clear cut
and subfields use the term
differently (see Appendix XXX).
%% Sentence deleted: point made already
% ARG inference was for many years synonymous [FIXME: this isn't really true,
% what about parsimony ARGs?]
% with estimating realisations
% of the coalescent with recombination (with, it must be said, limited success),
% and the distinction between process and realisation was not especially important.

@@

The word ``ARG'' is now widely used to refer to genetic genealogies in
general~\citep[MORE][]{mathieson2020ancestry}, including
varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
However, this broad usage means there is disagreement on the
exact object being defined. This causes problems, for example,
not just when evaluating inferred ARGs, but even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer}.
Perhaps most importantly, the direct linkage to
the original stochastic process definition leads to an overly
overly narrow interpretation of ARGs, which is significantly
out of step with modern methods, and
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we
assess success, and the expected limitations on the precision
and accuracy of our inferences.

% What do we do?
In this paper we attempt to address these problems by
providing a precise definition of an ARG that is general enough to encompass
the output of all popular simulation and inference methods.
% We follow the well established principle that it is advantageous
% to decouple the presence of mutations from the genealogical structure
% itself \citep[contra][chap.~3]{gusfield2014recombinatorics}.
We propose a definition of genetic genealogy in terms of specific genomes
and their intervals of genetic inheritance, and refer to this as
a ``genome ARG'', or ``gARG''.
This contrasts with the classical Griffiths approach,
based on recording evolutionary events, which we call
an ``event ARG'', or ``eARG''.
We show that the gARG encoding has a range of advantages over the
eARG approach, being flexible, efficient and, in particular,
allowing us to systematically encode uncertainty about the
time ordering of recombination events.

% We examineV}
% In particular, we argue that the goal of inferring fully precise
% ARGs---a necessity under the eARG encoding---should be
% % weird phrasing for a criticism?
% critically examined
% %
% as part of the process of fully decoupling our ARG data
% structure from any specific generative model.

@@

\section*{Event ARGs}\label{eARG}
In keeping with its original derivation in terms of a backward-time stochastic
process (see Appendix~\ref{sec-arg-history}),
the classical definition of an ARG data structure is in terms of evolutionary
\emph{events}.

and we therefore refer to it as an ``event ARG'' (eARG).
In an eARG, nodes represent
common ancestor or recombination events that occurred in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage with one parent, whereas in a recombination
event a lineage is split into two independent
ancestral lineages, and a list of two parents is required.
As illustrated in Figure~\ref{fig-arg-data-structure},
the regions of the genome which are transmitted are usually
specified by associating recombination nodes with a breakpoint.
In other words, eARGs traditionally
carry positional information on their nodes, rather than
storing genome positions as edge annotations.

@@

% \section*{Inheritance and ancestral material}\label{Ancestry_resolution}
\section*{Prospective vs retrospective ARGs}
ARGs are usually thought of as a retrospective structure, in which we
reason about the genomes that are genetic ancestors of some set of
sample genomes. Figure~\ref{fig-arg-in-pedigree}
is retrospective because it only includes information
about genetic inheritance relative to the sample individuals
$I_1$ and $I_2$ (genomes \textsf{a}--\textsf{d}).
Genetic transmission must have occurred between (e.g.)
$I_8$ and her daughter $I_6$, but this is omitted because
% this is probably not true, given we know the complete pedigree?
it is not relevant to the genetic ancestry of the samples
